### 2010-09-16 Thesis/Chapter2.tex
- Update 16 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@176 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\section{Tissue characterisation}
One of the main obstacles in developing realistic organ models is the lack of data on the material properties of live organ tissues. Measuring and characterising in vivo organ properties is a highly challenging task, but is a requirement for realistic organ modelling. Indeed, as we have seen in sections \ref{chap2:orthotropic} and \ref{chap2:non-linear}, constitutive equations require the experimental determination of material constants. Organ models with incorrect material properties will affect training in surgical simulator systems \citep{Sedef06}.

The constitutive equations are used to define idealised materials, which represent different aspects of the mechanical behaviour of natural materials. However, no real material is known to behave exactly as one of these mathematical models. In particular, few biological tissues obey Hooke's law. If because of their high stiffness (and therefore small deformations) cortical bones are accurately described by a linear model, Hooke's law becomes insufficient to describe the mechanical behaviour of the liver with precision for instance. In fact, in vivo experiments conducted by \cite{Melvin73} suggest that mechanical behaviour of liver is non-linear and viscoelastic. The brain also appears to be non-linear and viscoelastic \citep{Miller97}. The liver has the additional property of anisotropy, more precisely, it is transversely isotropic according to \cite{Chui07}.

These various mechanical aspects of materials make realistic modelling of anatomical structures fairly challenging. In medical simulation, where modelling must occur in real-time or close to real-time, very often simplifications have to be made. As we will see in chapter \ref{chap4} and \ref{chap7}, numerous models are available in the literature to describe physics of objects, from fairly simple and naive approaches to more complex and thorough representations. However, the most common technique to solve equations of continuum mechanics is the Finite Element Method. Moreover, since the two main contributions of this thesis deal with finite element methods formulations, the general principles of this method will be detailled in the next section.

>>>>

\section{Tissue characterisation}

\ON If continuum mechanics does teach us how to model idealised continuous structures, modelling soft tissues remains very complex. Indeed, in practice soft tissues are complex structures: they do not consist of a simple continuous piece of material, they are made of various constituents. Mechanical properties of soft tissues not only depends on the properties of these constituents, but how they are arranged relative to each other. This elaborate arrangement often leads to material properties for the global structure that are non-homogenous, non-linear, anisotropic and even viscoelastic.

As we know, \OFF the constitutive equations are used to define idealised materials, which represent different aspects of the mechanical behaviour of natural materials. \ON However, in practice, \OFF no real material is known to behave exactly as one of these mathematical models. In particular, few biological tissues obey Hooke's law. If because of their high stiffness (and therefore small deformations) cortical bones are accurately described by a linear model, Hooke's law becomes insufficient to describe the mechanical behaviour of the liver with precision for instance. In fact, in vivo experiments conducted by \cite{Melvin73} suggest that mechanical behaviour of liver is non-linear and viscoelastic. The brain also appears to be non-linear and viscoelastic \citep{Miller97}. The liver has the additional property of anisotropy, more precisely, it is transversely isotropic according to \cite{Chui07}.

These various mechanical aspects of materials make realistic modelling of anatomical structures fairly challenging. \ON The development of algorithms that allow the interactive computation of the deformation of soft tissues is very difficult. Therefore, \OFF in medical simulation, where modelling must occur in real-time or close to real-time, very often simplifications have to be made \ON and only the key characteristics of the material that are relevant for the objective of the simulation are considered and simulated. Experiments have to be performed to check the relevancy of the chosen constitutive model and to determine its parameters. At last, an optimised implementation of the model is required to perform in real-time.

\OFF One of the main obstacles in developing realistic organ models is the lack of data on the material properties of live organ tissues. Measuring and characterising in vivo organ properties is a highly challenging task, but is a requirement for realistic organ modelling. Indeed, as we have seen in sections \ref{chap2:orthotropic} and \ref{chap2:non-linear}, constitutive equations require the experimental determination of material constants. Organ models with incorrect material properties will affect training in surgical simulator systems \citep{Sedef06}. \ON Some devices exist for inert materials which allow the measure of Young's modulus, shear modulus and Poisson's ratio. However, the same measures are much more difficult to obtain for living tissues. In particular, experiments confirmed the important differences between in vitro and in vivo measures, as demonstrated by \cite{Kerdok06} with the viscoelastic properties of the liver for instance. Absence of blood perfusion and difference in temperature are two of most important factors to explain these differences. Obviously, in vivo constraints are very heavy and developing experiments and devices for in vivo measures of soft tissue is a very complex task. The preparation of samples and their testings (tensile, compression, thermal effects, rate of loading etc.) are not straightforward procedures.

A non-invasive technique called \emph{elastography} was introduced recently to measure Young's modulus. The first device based on this technique was commercialised by Echosens in 2003 \cite{Bastard09}. The techniques of elastography may be classified in two categories: static and dynamic elastography. In static elastography, a quasi-static constraint is applied onto the surface of the tissue and its deformation is measured. If the constraint is known, the Young's modulus may be obtained via Hooke's law. Dynamic elastography relies on shear waves propagation in the tissue. Shear waves are generated by an acoustic driver and the displacement caused by the waves is determined by ultrasounds, magnetic resonance imaging (MRI) or optical tomography. As an example, magnetic resonance elastography is a quantitative technique which allow the generation of maps describing the shear modulus within the tissues (see \fig{chap2:fig-elastography_liver}). However, elastography is still in development and only a few techniques of elastography are actually used clinically. In the future, elastography might be able to provide a map of the material properties of tissues which may be used as input data for any computational model. This technique would be a step towards patient-specific simulation. As diseased organs are often much stiffer than healthy structures, the integration of these data would greatly increase the fidelity of the simulation.
%
\begin{figure}
\centering
\includegraphics[width=12cm]{chapter2/elastography_liver.pdf}
\caption {\ON Magnetic resonance elastography of the liver in a healthy volunteer and a patient with cirrhosis. Image courtesy of \cite{Talwalkar07}.}
\label{chap2:fig-elastography_liver}
\end{figure}

Yet, precised characterisation of soft tissue is not sufficient to achieve a good simulation. The development of a simplified or optimised version of the model suitable for real-time computation is also mandatory. \OFF As we will see in chapter \ref{chap4} and \ref{chap7}, numerous models are available in the literature to describe physics of objects, from fairly simple and naive approaches to more complex and thorough representations. However, the most common technique to solve equations of continuum mechanics is the finite element method. Moreover, since the two main contributions of this thesis deal with finite element methods formulations, the general principles of this method will be detailed in the next section.

### 2010-09-16 Thesis/Chapter4.tex
- Update 16 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@176 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\TODO{add drawbacks of meshless methods: supposedly good for handling topological changes but other issues. Couldn't really find drawbacks in meshless' literature.}

>>>>

\ON The main advantage of meshless methods is of course in the absence of a mesh. This advantage of not having to go through the complex process of meshing was often claimed by meshless methods partisans. It also allows much easier cutting as it saves the cost of a computationally demanding remeshing around the cut. Yet, in contrast of finite element methods, meshless techniques are young and still in development. Boundary conditions are an aspect of the numerical solution of partial equations where meshless methods have many difficulties for instance. According to \cite{Horton10}, if meshless methods are accurate in terms of overall reaction forces, they are not as good with individual displacements or forces. Therefore the authors suggest to use a meshless algorithm to fill the interior sections of an organ (where a local discrepancy has minor consequences) and use the finite element method for the boundary. As a conclusion, if meshless methods are promising, the technique does not seem as mature as the finite element method for modelling soft tissues. 
\OFF
